A Simple Minibatch Acceptance Test for MCMC

### 2016-05-17 nips_2016.tex
Updated the Algorithm 1 portion of the NIPS draft

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
If we fix $\theta_t$ and $\theta'$, the randomness in our $\Delta'$ samples only comes from the
minibatch. Since the minibatch is chosen randomly, the distribution of $\sum_{i=1}^n (\log p(x_i\mid
\theta') - \log p(x_i\mid \theta_t))$ converge to a Gaussian by the Central Limit Theorem.
\end{proof}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship can be expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise in $\Delta'$ means we can no longer perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose
$X$ as $X = X_{\rm norm}+X_{\rm corr}$, where the first term is a zero-mean ``noise term'' which
has the same variance as $\varepsilon$, and the second term is a zero-mean ``correction'' term.
These two convolve to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
This deconvolution process is visualized in Figure~\ref{fig:part2}. {\color{blue} (Daniel: note that
we will be fixing this figure, so the visualization should hopefully be a little clearer.)} We
decompose $X$ this way so that we can estimate the two components separately.  Notice that if we
increase the mini-batch size, this will cause the $X_{\rm noise}$ distribution to shrink and become
narrower.

Furthermore, since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
inequality in Equation~\ref{eq:criteria} an equality.

Algorithm~\ref{alg:our_algorithm} concisely describes our MH test.

Our mini-batch MH test can not only be used to perform MCMC, but also can be used to obtain parameter samples around MAP/ML by adding temperature $t$ to the likelihood function. Thus, the target distribution become $f = \left ( \prod_{i=1}^N p(x_i \mid \theta') \right )^{t_i}$. We can decrease the temperature at each iteration to decline the mixing rate, so that our samples will converge to the MAP/ML point.

{\color{blue}
Daniel: I really need to make this clear and do the algorithm description.
}

\begin{comment}
though it really should be an approximation as we can't guarantee that our $X_{\rm norm}$ term is
exactly the correct amount of noise), where the left hand side has distributions representing normal
noise ($X_{\rm norm}$, which is labeled as ${\rm Noise}_U$, but they are the same thing) and a
``correction'' term labeled $X_{\rm correction}$. Be aware that all of these ``additions'' are to be
interpreted as convolutions when computing the sum of densities.

I think it's becoming clear but I'm still trying to connect all of the steps.  During each iteration
of MCMC, we compute $\Delta'$ and \emph{sample} (right?) an $X$ (which we can do by sampling $u$ and
then doing $X = g^{-1}(u)$. This gives us $\Delta' + X$ but our MH test actually requires $X_{\rm
correction}$, so first we must estimate $\sigma^2$, the variance of $X_{\rm noise} \sim
\mathcal{N}(0, \sigma^2)$ (perhaps from the mini-batch data, somehow?) and then somehow ``remove''
that component from $X$ to get $X_{\rm correction}$? Then we have our acceptance test. Is this
right? Or, of course, we don't have to sample $X$ at all, but just figure out a way to sample
$X_{\rm correction}$ directly.

I think we talked about estimating $\sigma^2$. The variance $\sigma^2$ that we're concerned about is
a scalar value based on the sum of log probabilities of different samples, so one way we could
estimate the variance is by taking all samples of a mini-batch and seeing how much those individual
$\log p(x_i \mid \theta)$ vary. Or we could look at $k$ mini-batches of data which will give us
$\Delta_1, \ldots, \Delta_k$ and compute the empirical $\sigma^2$ (of course that defeats the
purpose of using mini-batches).

I agree with you that out MH test requires $X_{correction}$. My understanding
is that we can actually estimate $X$ from the whole data set, and the probability distribution of
$X_{norm}$ is what we proposed? So the determination of $X_{correction}$ is a deterministic
deconvolution problem.
\end{comment}

\begin{algorithm}[t]
\KwData{this text}
\KwResult{how to write algorithm with \LaTeX2e } initialization\;
\While{not at end of this document}{
    read current\;
    \eIf{understand}{
        go to next section\;
        current section becomes this one\;
    }{
        go back to the beginning of current section\;
    }
}
\caption{A description of our MH test.}
\label{alg:our_algorithm}
\end{algorithm}

>>>>

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian by the
Central Limit Theorem.
\end{proof}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship can be expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise in $\Delta'$ means we can no longer perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose
$X$ as $X = X_{\rm norm}+X_{\rm corr}$, where the first term is a zero-mean ``noise term'' which
has the same variance as $\varepsilon$, and the second term is a zero-mean ``correction'' term.
These two convolve to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
This deconvolution process is visualized in Figure~\ref{fig:part2}. We decompose $X$ this way so
that we can estimate $X_{\rm norm}$ from the data, and then add the correction $X_{\rm corr}$ later.
Notice that we have some control over the true value of $X_{\rm norm}$, even though we cannot
generally guarantee computing it exactly. In particular, increasing the minibatch size will cause
the $X_{\rm norm}$ distribution to become narrower. {\color{blue} (Daniel: note that we will be
fixing this figure, so the visualization should hopefully be a little clearer, along with the
corresponding paragraph description. In particular, this paragraph needs an informal description of
Algorithm~\ref{alg:our_algorithm}.)} Algorithm~\ref{alg:our_algorithm} concisely describes our MH
test.

Since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
inequality in Equation~\ref{eq:criteria} an equality.

It is also important to note that our minibatch MH test can also be used to obtain parameter
estimates close to the MAP and/or ML estimaes by augmenting the likelihood function with a
temperature term $T > 1$. Thus, the target distribution would become
\begin{equation}\label{eq:temperature}
f(\theta') = p(\theta')\left ( \prod_{i=1}^N p(x_i \mid \theta') \right)^{1/T}.
\end{equation}
We can decrease the temperature as the MCMC sampling proceeds, thus slowing the mixing rate and
sharpening the peaks of the distribution, so that the samples coverge to the desired point estimate.

{\color{blue}
Daniel: Haoyu and I are going to make this section a lot more clearer, along with the algorithm
description.
}

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

\begin{algorithm}[t]
\Input{Samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm corr}$ values.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or point estimation.}
Initialize $\theta_1$\;
\For{$t=\{2, \ldots, T\}$}{
    Create a new empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, K\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation of the $\Delta'$ elements from list $d$\;
    Using a random minibatch of $m$ data points, compute a \emph{new} $\Delta'$; call it $\Delta_{\rm real}'$\;
    Sample $u \sim {\rm Unif}[0,1]$ and extract the corresponding $X_{\rm corr}$ from pre-computed data\;
    Re-scale the correction error: $X_{\rm corr} = X_{\rm corr} \cdot {\rm std}(d)$\;
    {\color{blue} Daniel: TODO, check if we need some variance correction or something here.}\;
    \eIf{$X_{\rm corr} + \Delta_{\rm real}' > 0$}{
        Accept the candidate, $\theta_{t+1} = \theta'$\;
    }{
        Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
    }
}
\caption{A description of our MH test. {\color{blue} Daniel: IN PROGRESS!}}
\label{alg:our_algorithm}
\end{algorithm}
